---
title: "Creating Maps with ggplot2"
---

## Introduction

In this exercise we will use [_simple features_](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html) and
_ggplot2_ to create a map of Ethiopia and then a map of two regions highlighting the regions within a larger map of Ethiopia.
Once we have the base maps of the country and regions we will add points and zones to the map to indicate survey sites.

```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(ggsn)
library(readr)
library(sf)
library(stringr)
```

## Create Country Level Map of Ethiopia

Several packages offer direct access to map layers in R for use in creating maps.
However, in this exercise we will download the data directly from <https://www.gadm.org>.
GADM is a free resource to use for non-commercial and research purposes that provides data for mapping administrative boundaries.
You can find out more at <https://gadm.org/about.html>.

For our map we will download an _sf_ object.
_sf_ stands for "Simple Features", which means that geospatial data is represented as records in a `data.frame` in R, making it easier to work with than previous spatial data formats for R.
_sf_ plays nicely with the _tidyverse_ making it ideal for using with _ggplot2_ when making maps.

### Using download.file() to Fetch Data

Using `download.file()`, we will download an _sf_ file from the GADM website.
`download.file()` is a base R function useful for retrieving data from a web URL and making it available in your R session.
Here we will download the data to R's `tempdir()`, a temporary directory that R uses during your session.
***Nothing here is saved after you exit R.***
If you wish to save the GADM file, you should specify another location, _e.g._ in your Documents or Home directories to save the file for later use, you can do this in the 
Using this method allows us to all download the file and import it into R using the same commands without worrying about usernames, operating systems, etc.

This next section provides the URL to the .rds file (a native R file format) containing the data for mapping Ethiopia.
Then we provide the file path to save the file in the `tempdir()`.
Next, we check what operating system is being used to download the data.
Windows handles the downloads differently and will corrupt the file if not properly handled.
The `if()` section checks the OS and then proceeds to download the file using the proper method depending on the OS detected.
Lastly using `readRDS()` we import the data and create our first plot of Ethiopia!

```{r download-GADM, message=FALSE, warning=FALSE}
ethiopia_rds <-
  "https://biogeo.ucdavis.edu/data/gadm3.6/Rsf/gadm36_ETH_3_sf.rds"
ethiopia_sf <- file.path(tempdir(), "gadm36_ETH_3_sf.rds")

if (toupper(Sys.info()["sysname"]) == "WINDOWS") {
  download.file(
    url = ethiopia_rds,
    destfile = ethiopia_sf,
    method = "wininet",
    mode = "wb"
  )
} else {
  download.file(ethiopia_rds, ethiopia_sf, method = "auto")
}

ethiopia_sf <- readRDS(ethiopia_sf)
```

Creating maps using _ggplot2_ and _sf_ follows the same principals that Felipe has already demonstrated with `geom_*()`s, layers, themes, colours and fills.

First, plot the whole country using _ggplot2_'s `geom_sf()` and the `ethiopia_sf` object.

```{r plot_level1}
ethiopia_ggplot <- ggplot(ethiopia_sf) +
  geom_sf(fill = "white")

ethiopia_ggplot
```

We can see that there are several zones, shown as the smaller polygons within the larger country outline, in this data set.

### Add Labels to the Regions

The GADM data have several columns to work with.
You can find out what these are by using the `names()`.

```{r colnames}
names(ethiopia_sf)
```

We can group the GADM data using `dplyr::group_by()` to group the polygons into larger regions and add labels to them.
In this data set, there are three levels and three columns we might be interested in.

* `NAME_0` refers to the country level and gives a whole country outline,
* `NAME_1` refers to the second level of administrative boundaries, _e.g._ states within the USA,
* `NAME_2` refers to the third level of administrative boundaries, _e.g._ counties or parishes within a US state, and
* `NAME_3` refers to the fourth level of administrative boundaries, _e.g._ districts within counties in a US state.

Grouping by any value, 0 - 2 will aggregate or dissolve the lines within the map producing larger polygons.
Here we will group by `NAME_1` for the regions.

```{r group-NAME_1, warning=FALSE, message=FALSE}
ethiopia_regions_sf <-
  ethiopia_sf %>% 
  group_by(NAME_1) %>%
  summarise() %>%
  ungroup() %>%
  st_as_sf()
```

Using `geom_sf_text()` we can add labels to the map.

```{r plot-NAME_1, warning=FALSE, message=FALSE}
ethiopia_regions_ggplot <- 
  ggplot(ethiopia_regions_sf) +
  geom_sf(fill = "white") +
  geom_sf_text(
    aes(label = NAME_1),
    size = 2.5,
    hjust = 1
  )

ethiopia_regions_ggplot
```

The _ggplot2_ theme, `theme_bw()` is nice to use with maps, so we will apply that here as well as labelling the x and y axis with "longitude" and "latitude" labels.

```{r theming-labelling, warning=FALSE, message=FALSE}
ethiopia_regions_ggplot <- 
  ethiopia_regions_ggplot +
  labs(x = "Longitude",
       y = "Latitude") +
  theme_bw()

ethiopia_regions_ggplot
```

## Adding and Labelling Points on a Map

Now we will build on the maps we created by adding the sampling location points that Del Ponte and Belachew (2020) used with latitude and longitude values from their CSV file.

### Create data.frame of Sampling Locations

Using the previously created `ethiopia_regions_sf` object, we will plot the locations of the survey sites from Del Ponte and Belachew (2020).
We will download their CSV directly into our R session from the GitHub repository.

```{r coffee-sampling, message=FALSE}
coffee_sampling <-
  read_csv(
    "https://raw.githubusercontent.com/emdelponte/paper-coffee-rust-Ethiopia/master/data/survey_clean.csv"
  )
```

### Converting Sites to an _sf_ Object

While we could plot everything using `geom_point()` directly from the .CSV file `data.frame`, but if we convert the `data.frame` to an `sf` type object it is much more flexible for mapping with other _sf_ objects in _ggplot2_.
The conversion is done using `st_as_sf()` and providing the columns that hold the coordinates (latitude and longitude).
The `crs` argument tells _sf_ what projection we want to use.
Map projections are used to deal with the distortion of creating a flat figure to represent a spherical surface.
In this case we'll use EPSG:4326, which is only latitude and longitude values.
While we won't go into projections here, the [_sf_ vignette](https://r-spatial.github.io/sf/articles/sf5.html#how-does-sf-project-geographic-coordinates) describes some of the ideas behind this and many more web resources exist to help you pick the right projection for your map.

```{r convert-coffee-to-sf, warning=FALSE, message=FALSE}
coffee_sampling_sf <-
  st_as_sf(coffee_sampling,
           coords = c("lon", "lat"),
           crs = 4326)
```

Now we're ready to add the new _sf_ object of sampling points to the map!

### Adding Sampling Locations to Map

Adding the points to the map works in the same fashion as adding layers in other _ggplot2_ objects.

```{r add-points, warning=FALSE, message=FALSE}
ggplot() +
  geom_sf(data = ethiopia_regions_sf,
          fill = "white") +
  geom_sf(data = coffee_sampling_sf,
          size = 2,
          color = alpha("black", 0.35)) +
  labs(
    title = "Ethiopian Coffee Leaf Rust Survey Sites",
    caption = "Data from GADM, gadm.org, and Del Ponte & Belachew (2020)
    https://doi.org/10.17605/OSF.IO/XEJAZ",
    fill = "Zone",
    linetype = "Region",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw()
```

## Highlighting Regions

In some cases it is desirable to create maps highlighting certain areas, _e.g._ states, regions, counties, etc.

Del Ponte and Belachew (2020) sampled in two regions of Ethiopia, Oromia and Southern Nations, Nationalities and Peoples' Region (SNNP).

To create a map that highlights Oromia and SNNP, we will dissolve internal borders leaving only the region level borders using `group_by()` and `summarise()`.
For simplicity and labelling the map, we will replace Southern Nations, Nationalities and Peoples" with "SNNP".
Lastly, `ungroup()` and convert the resulting `tibble()` back to an _sf_ object.

### Dissolving Internal Boundaries for Region Maps

```{r subset-snnp-oromia, warning=FALSE, message=FALSE}
ethiopia_regions_sf <-
  ethiopia_regions_sf %>% 
  mutate(NAME_1 = gsub("Southern Nations, Nationalities and Peoples", "SNNP",
                       NAME_1)) %>%
  group_by(NAME_1) %>%
  summarise() %>%
  ungroup() %>%
  st_as_sf()

ethiopia_regions_sf
```

After grouping by `NAME_1`, we can map the new object and see how it compares to the original data with finer subnational divisions.

```{r ggplot-ssnp-oromia, warning=FALSE, message=FALSE}
ggplot(data = ethiopia_regions_sf) +
  geom_sf() +
  labs(x = "Longitude",
       y = "Latitude") +
  theme_bw()
```

### Highlighting Specific Regions Within the Country

If you wish to highlight the regions, calling them out on the map as Del Ponte and Belechew (2020) did in their work, we can create a new object of only these two regions and plot them in a new layer.

To do this we'll create two new layers to work with.
The first layer will just be Ethiopia with no regions, zones or other subdivisions.
The second layer will be the layer that contains Oromia and SNNP.

```{r filter-ssnp-oromia, warning=FALSE, message=FALSE}
# First layer, country outline
ethiopia_simple_sf <-
   ethiopia_sf %>%
   group_by(NAME_0) %>%
   summarise() %>%
   ungroup() %>%
  st_as_sf()

# Second layer, regions we're interested in
oromia_snnp_sf <- filter(
  ethiopia_regions_sf,
  NAME_1 == "Oromia" |
    NAME_1 == "SNNP"
)
```

Plot the complete map.

```{r plot ssnp-oromia}
ggplot() +
  geom_sf(data = ethiopia_simple_sf,
          col = NA) +
  geom_sf(data = oromia_snnp_sf,
          aes(linetype = NAME_1,
              fill = NAME_1)) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    caption = "Data from GADM (gadm.org).",
    fill = "Region",
    linetype = "Region",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw()
```

## Going Further

Del Ponte and Belechew (2020) display a map that highlights the two regions, Oromia and SNNP with the zones within them where surveys were conducted also coloured.
We can recreate this map, but will do so without the labels that they used.
Here we will only plot the points that were surveyed as before on top of the two regions and nine zones surveyed.

First create another new layer of only the zones that were involved in the survey.
Then construct the map using the layers we've already created.
Lastly, using functions from _ggsn_, `scalebar()` and `north()`, add a scale bar to the bottom right and a north arrow to the upper right to orient viewers who may not be familiar with Ethiopia.

```{r filter-zones, warning=FALSE, message=FALSE}
ethiopia_zones_sf <- 
  ethiopia_sf %>%
  filter(
    NAME_2 %in% c(
      "Jimma",
      "Mirab Welega",
      "Sidama",
      "Sheka",
      "Keffa",
      "Bench Maji",
      "Bale",
      "Gedeo",
      "Ilubabor"
    )
  )
```

Plot the data to check.

```{r final-map}
final_map <- ggplot() +
  geom_sf(data = ethiopia_simple_sf,
          col = NA,
          fill = "#D5C1AB") + # use a coffee-ish colour background for country
  geom_sf(data = oromia_snnp_sf, # add the two regions in white
          aes(linetype = NAME_1),
          fill = "white") +
  geom_sf(data = ethiopia_zones_sf, # add zones and fill by name
          aes(fill = NAME_2),
          colour = NA) + # no outline colour
  geom_sf(data = coffee_sampling_sf,
          # add sampling points
          size = 2,
          colour = alpha("black", 0.35))

final_map
```

Add linetype definitions for regions and add zone colours.
The Colorbrewer palette offers good colours for maps, see <https://colorbrewer2.org> for more ideas.
Here we'll use `scale_fill_brewer()`, a built in _ggplot2_ colour scale that can be used with any type of plot, not just maps.

```{r final-map-legend}
final_map <- 
  final_map +
  scale_linetype( # define region outline linetype by region
    labels = function(x)
      str_wrap(x, width = 5)
  ) +
  scale_fill_brewer(palette = "Set3") + # this colours the zones, https://colorbrewer2.org/
  xlab("Longitude") +
  ylab("Latitude") +
  labs(
    title = "Ethiopian Coffee Leaf Rust Survey Sites",
    caption = "Data from GADM, gadm.org, and Del Ponte & Belachew (2020)
    https://doi.org/10.17605/OSF.IO/XEJAZ",
    fill = "Zone",
    linetype = "Region",
    colour = ""
  )

final_map
```

Lastly, add standard map items like a scale bar and north arrow using functions from _ggsn_, `north()` and `scalebar()`.

```{r final-touches}
final_map <-
  final_map +
  north(ethiopia_simple_sf) + # from ggsn
  scalebar( # from ggsn
    ethiopia_simple_sf,
    dist = 250,
    st.size = 2.5,
    dist_unit = "km",
    transform = TRUE,
    model = "WGS84"
  ) +
theme_bw()

final_map
```

The final map includes a title, a data acknowledgement and clearly labelled legend describing the contents.
Aside from using _ggsn_ functions, the map is generated using familiar _ggplot2_ commands.

## Further Reading and Resources

[Drawing beautiful maps programmatically with R, sf and ggplot2 — Part 1: Basics](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html)

[Drawing beautiful maps programmatically with R, sf and ggplot2 — Part 2: Layers](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf-2.html)

[Drawing beautiful maps programmatically with R, sf and ggplot2 — Part 3: Layouts](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf-3.html)
