---
title: "Visualizing Networks"
author: "Kelsey Andersen, Robin Choudhury"
---

For this portion of the workshop, we will use the following packages:

```{r, results='hide', message=FALSE}
library(igraph)
library(ggplot2)
```

# Networks with igraph 

## Simple Networks

First, let's create a simple adjacency matrix with three rows and three columns: 
```{r}
mat1 <- matrix(c(0, 1, 0, 0, 0, 1, 1,0, 0), nrow=3, ncol=3) ### matrix function 
mat1
```


Use the igraph function *graph_from_adjacency_matrix()* to create a network object from your graph, then use the *plot()* function to plot. Use edge.arrow.size to set the size of the arrows. 
```{r}
mat2 <- graph_from_adjacency_matrix(mat1)
plot(mat2, edge.arrow.size = 1) ## set the size of the arrows 
```
\  

Alternatively, create the same network by telling igraph what links you would like and how many nodes you would like to establish. We also want to tell igraph if our network is directed (with arrows) or undirected. 

```{r}
mat3 <- graph(edges=c(1,3, 3,2, 2,1), n=3, directed=T ) # use graph function and list edges 
plot(mat3, edge.arrow.size = 1)
```
\  
# Network Aestetics  
Many parts of a network can be sized and colored to help communicate results more clearly.  
Here, for example, we color the nodes and change the size and position of the labels using *vertex.color =* and *vertex.label.dist =*

```{r}
plot(mat3, edge.arrow.size = 1, vertex.color = "purple", vertex.size = 25, vertex.label.cex = 1)

```

## Labels

Here we use *vertex.label.cex* to make our labels slightly bigger, and we use *vertex.label.dist* to move the labels off the network. This can be useful, particularly when we have large networks. 

```{r}
plot(mat3, edge.arrow.size = 1, vertex.color = "purple", vertex.size = 25, vertex.label.cex = 2, vertex.label.dist = 3.5)
```

## Curved edges

Sometimes it is useful to curve edges of our network: 

```{r}
plot(mat3, edge.arrow.size = 1, vertex.color = "purple", vertex.size = 25, vertex.label.cex = 2, vertex.label.dist = 3.5, edge.curved = .2)
```

## Colors

Here we color the node borders the same color as the nodes themselves, but this could be modified to a different color that gives additional information about the node: 

```{r}
plot(mat3, 
     edge.arrow.size = 1, 
     vertex.color = "purple", 
     vertex.frame.color="purple",
     vertex.size = 25, 
     vertex.label.cex = 2, 
     vertex.label.dist = 3.5, 
     edge.curved = .2)
```

We can also change edge colors. This can be useful when edges mean different things. 

```{r}
plot(mat3, 
     edge.arrow.size = 1, 
     vertex.color = "purple", 
     vertex.frame.color="purple",
     vertex.size = 25, 
     vertex.label.cex = 2, 
     vertex.label.dist = 3.5, 
     edge.curved = .2,
     edge.color="purple")
```

## Other ways to change aesthetics

We can also adjust plot aesthetics by applying changes directly to the verticies *V()* or edges *E()* of our plots. 

```{r}
V(mat3)$color <- 'purple'
V(mat3)$size <- 25
E(mat3)$color <- 'purple'
E(mat3)$arrow.size <- 1
V(mat3)$frame.color <- 'purple'
plot(mat3)
```


# Network data types  

## Adjacency matrices vs node lists
\  
You can read in your data directly as an adjacency matrix, but likely this is not the way that you have your data organized. Instead, it might be easier to have two files: a *node* file and an *edge* file.  

In a node file, the first two columns are all of your from:to links. Column 1 is always *from*, Column 2 is always *to* (less important for undirected networks). The columns after that are your edge attributes (such as weight of link, volume, probability, name etc).  
\  

Here is an example of a simple node list, where all of the nodes are farmers and links are communication. We include attributes about the *node* like age, gender, and number of years farming.  
```{r}
Nodelist <- data.frame(
                    Names =c("Jim", "Carole", "Joe", "Michelle", "Jen", "Pete", "Paul", "Tim", 
                            "Jess", "Mark", "Jill", "Cam", "Kate") ,
                    YearsFarming = c(8.5, 6.5, 4, 1, 3, 10, 5, 5, 5, 1, 1, 6, 6) , 
                    Age = c(27, 52, 49, 32, 65, 72, 42, 67, 48, 33, 67, 75, 39) , 
                    Gender = c("Male", "Female", "Male", "Female", "Female", "Male","Male","Male", "Female", "Male", "Female", "Male", "Female"))
Nodelist       
```
\  

Now we can create an edgelist data frame- Who shared information in the 2019 growing season? How frequently? 
```{r}
Edgelist <- data.frame(
                  From = c("Jim", "Jim", "Jim", "Jill", "Kate", "Pete", "Pete", "Jess", "Jim", "Jim", "Pete"),
                  To = c("Carole", "Jen", "Pete", "Carole", "Joe", "Carole", "Paul", "Mark", "Cam", "Mark", "Tim"),
                  Times =c(3, 7, 6, 6, 5, 3, 2, 1, 1, 2, 5)
)

```


# igraph objects

Let's make our farmer communication network!
```{r}
FarmNetwork <- graph_from_data_frame(d = Edgelist, vertices = Nodelist, directed = T)
FarmNetwork
E(FarmNetwork) # view edges
V(FarmNetwork) # view nodes
```


Plot!
```{r}
plot(FarmNetwork, edge.arrow.size = .5, vertex.color = "plum", vertex.label.dist = 2.5)

```

## Node colors
 Much more information about making beautiful networks in R using igraph can be found at [Katya Ognyanova's Site](http://kateto.net/networks-r-igraph). But briefly:  
\  
Let's color our nodes based on gender
```{r}
V(FarmNetwork)$color <- ifelse(V(FarmNetwork)$Gender == "Male", "#4682B4", "#C83200")
V(FarmNetwork)$frame.color <- ifelse(V(FarmNetwork)$Gender == "Male", "#4682B4", "#C83200")
plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5, edge.curved = .2)

```
## Layouts 

You will notice that evertime that you run the above code, the layout changes. Here we will show you layout options, and how to "fix" a layout once you have chosen one that you like. 
```{r}
Layout1 <- layout_in_circle(FarmNetwork)
plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5, layout = Layout1)
```

```{r}
Layout2 <- layout_as_star(FarmNetwork)
plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5, layout = Layout2)
```


Layout based on the Kamada-Kawai layout algorithm which places the nodes on a plan based on a physical model of springs. 
```{r}
Layout3 <- layout_with_kk(FarmNetwork)
plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5, layout = Layout3)
```


```{r}
Layout4<- layout_with_fr(FarmNetwork)
plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5, layout = Layout4)
```

\  
You can also size your nodes based on attributes:
```{r}

V(FarmNetwork)$size <- V(FarmNetwork)$YearsFarming # size the nodes by number of years farming
plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5)

```

## Node size
Scale the node size up a bit.. 
```{r}

V(FarmNetwork)$size <- V(FarmNetwork)$YearsFarming *2 ## scale by multiplying by 2
plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5)

```

We can also adjust the size of the links. Here we will size them based on the number of times any two farmers communicated. 

## Link size

```{r}
E(FarmNetwork)$width <- E(FarmNetwork)$Times/1.5
plot(FarmNetwork, edge.arrow.size = .7, vertex.label.dist = 2.5, edge.curved = .2, layout = Layout2)
```

## Legends

Finally, let's add a legend so we understand what our colors mean. 

```{r}
plot(FarmNetwork, edge.arrow.size = .7, vertex.label.dist = 2.5, edge.curved = .2, layout = Layout1)
legend(x=-1.5, y=-1.1, c("Male","Female"), pch=21, pt.bg=c("#4682B4", "#C83200"), pt.cex=2, cex=.8, bty="n", ncol=1)

```

## Saving 

We can save our networks in a similar was as our ggplots, using ggsave. The only change that we need to make is to "wrap" ggsave around our final network. 
```{r}
ggsave(plot(FarmNetwork, edge.arrow.size = .5, vertex.label.dist = 2.5),
       filename = "graph/Network.tiff", width =150, height = 150, units = "mm",
         dpi = 300)
```



